Forecasting the Changing Dynamics of US Energy Product Exports to OECD Markets: The Role of the Russian-Ukraine Conflict and Future Scenarios

### 2023-05-22 Final Report_FM.Rmd
updated

In this study, we will analyze the impact of the war in Ukraine on US energy exports to OECD Europe. Specifically, we will focus on three main energy products: crude oil, petroleum products, and natural gas. By examining trends in these exports, we can better understand how the geopolitical events in Ukraine have affected the energy market in Europe. Additionally, we will investigate changes in OECD Europe's dependence on Russian gas to see how energy dynamics in the region have shifted.   
\linebreak  
\linebreak     
## 2. Time Series Graphics

@@

@@

@@

```{r, data_prep, echo=TRUE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray'}
library(fpp3)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(scales)

@@

```
**Comment on Nat Gas export zeros**

### 2.2 Time Series Plots

Firstly, we will examine a general overview of the three products: US exports of Crude Oil, Petroleum Products, and Natural Gas to OECD countries in barrels over the last ten years. This analysis will provide us with an understanding of the overall trend and magnitude of the exports. By examining the trend, we can identify any significant changes in the exports over the years and evaluate the magnitude to determine which product has the highest export value. This overview will be a crucial step before we dive deeper into the individual time series of each product.

### 2.3 Including Plots

The graph below reveals significant growth in Crude oil and Natural gas exports since around 2015, with a clear year-on-year increase in both goods, indicating a noticeable trend. Moreover, it appears that the time series for these two products are non-stationary, possibly suggesting the presence of a trend, seasonality, or both. In 2022/2023, Crude oil and Natural gas exports reached their peak, with crude oil almost 60,000 and natural gas around 40,000. Conversely, the time series for Petroleum products appears to be stationary, with a slight increasing trend that may require further analysis.

```{r,Times series overivew, echo=TRUE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray'}

@@

In the graph below, we have plotted each exported product separately to provide a better picture of each time series. The data shows that crude oil exports began in 2016 with lower amounts initially, but within a few years, exports boomed. Since 2016, there has been a significant positive trend, possibly with seasonality. In 2020, crude oil exports reached a peak of around 50,000 barrels. Then, due to the COVID-19 pandemic, exports decreased for two years. However, in the beginning of 2022, exports increased significantly, most likely due to the Russian invasion in Ukraine. Currently, exports of crude oil are at a peak, reaching almost 60,000 barrels.

Looking at natural gas exports, we can see that the US did not export to OECD countries until after 2016, when low amounts of natural gas were first exported. In 2018, exports increased significantly, reaching their first peak in 2020. We can see a similar positive trend and possible seasonality as in the crude oil graph. During the COVID-19 pandemic, exports decreased significantly, but again, after the Russian invasion, exports increased significantly, reaching 40,000 barrels. This suggests that OECD countries stopped buying natural gas from Russia.

Finally, looking at petroleum product exports, we can see some fluctuations since 2010, but no clear trend. Seasonality may exist. The changes during COVID-19 and after the Russian invasion were not as significant as in the other two graphs.

```{r, overview by product, echo=TRUE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray'}

@@

```
#### 2.4 Time Series Patterns

@@

@@

The Box-Cox transformation was already applied to stabilize the variance in chapter [3.2. Transformations]. Other than that no transformations were applied on neither of time series.

Following a definition of stanioarity provided by Rob J Hyndman and George Athanasopoulos in Forecasting: Principles and Practice (3rd ed) **("A stationary time series is one whose statistical properties do not depend on the time at which the series is observed")** it is obvious that our time series is non stationary since it has clear trend and some seasonality.

We start by performing Unit Root test to identify which methods are suitable in our case
```{r, Unit root test}

@@

```{r, first order difference}

@@

```{r}

@@

@@

>>>>

In this study, we will analyze the impact of the war in Ukraine on US energy exports to OECD Europe. Specifically, we will focus on three main energy products: crude oil, petroleum products, and natural gas. By examining trends in these exports, we can better understand how the geopolitical events in Ukraine have affected the energy market in Europe. Additionally, we will investigate changes in OECD Europe's dependence on Russian gas to see how energy dynamics in the region have shifted.   
\linebreak  
\linebreak     
## 2. Time Series Graphics      
\linebreak     

@@

\linebreak     
\linebreak     

@@

\linebreak     

@@

```{r, data_prep, echo=TRUE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray'}
library(fpp3)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(scales)

@@

```
\linebreak     
\linebreak     
### 2.2 Time Series Plots   
\linebreak     
Firstly, we will examine a general overview of the three products: US exports of Crude Oil, Petroleum Products, and Natural Gas to OECD countries in barrels over the last ten years. This analysis will provide us with an understanding of the overall trend and magnitude of the exports. By examining the trend, we can identify any significant changes in the exports over the years and evaluate the magnitude to determine which product has the highest export value. This overview will be a crucial step before we dive deeper into the individual time series of each product.  
\linebreak 
\linebreak     
### 2.3 Including Plots   
\linebreak     
The graph below reveals significant growth in Crude oil and Natural gas exports since around 2015, with a clear year-on-year increase in both goods, indicating a noticeable trend. Moreover, it appears that the time series for these two products are non-stationary, possibly suggesting the presence of a trend, seasonality, or both. In 2022/2023, Crude oil and Natural gas exports reached their peak, with crude oil almost 60,000 and natural gas around 40,000. Conversely, the time series for Petroleum products appears to be stationary, with a slight increasing trend that may require further analysis.  
\linebreak     
```{r,Times series overivew, echo=TRUE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray'}

@@

```
\linebreak     
In the graph below, we have plotted each exported product separately to provide a better picture of each time series. The data shows that crude oil exports began in 2016 with lower amounts initially, but within a few years, exports boomed. Since 2016, there has been a significant positive trend, possibly with seasonality. In 2020, crude oil exports reached a peak of around 50,000 barrels. Then, due to the COVID-19 pandemic, exports decreased for two years. However, in the beginning of 2022, exports increased significantly, most likely due to the Russian invasion in Ukraine. Currently, exports of crude oil are at a peak, reaching almost 60,000 barrels.  
\linebreak     
Looking at natural gas exports, we can see that the US did not export to OECD countries until after 2016, when low amounts of natural gas were first exported. In 2018, exports increased significantly, reaching their first peak in 2020. We can see a similar positive trend and possible seasonality as in the crude oil graph. During the COVID-19 pandemic, exports decreased significantly, but again, after the Russian invasion, exports increased significantly, reaching 40,000 barrels. This suggests that OECD countries stopped buying natural gas from Russia.  
\linebreak     
Finally, looking at petroleum product exports, we can see some fluctuations since 2010, but no clear trend. Seasonality may exist. The changes during COVID-19 and after the Russian invasion were not as significant as in the other two graphs.  
\linebreak     
```{r, overview by product, echo=TRUE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray'}

@@

```
\linebreak     
\linebreak

#### 2.4 Time Series Patterns

@@

**Komentaras apie bendra triju eksportu ir priklausomybes tendencija**

@@

The Box-Cox transformation was already applied to stabilize the variance in chapter **[3.2. Transformations]**. Other than that no transformations were applied on neither of time series.

Following a definition of stanioarity provided by Rob J Hyndman and George Athanasopoulos in Forecasting: Principles and Practice (3rd ed) *("A stationary time series is one whose statistical properties do not depend on the time at which the series is observed")* it is obvious that our time series is non stationary since it has clear trend and some seasonality.

We start by performing Unit Root test to identify which methods are suitable in our case
```{r, Unit root test, echo=FALSE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray',echo=FALSE}

@@

```{r, first order difference,echo=FALSE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray',echo=FALSE}

@@

```{r stationarity, echo=FALSE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray',echo=FALSE}

library(ggplot2)
library(dplyr)
library(tsibble)
library(lubridate)
library(forecast)

@@

```
```{r, ARIMA estimation, echo=FALSE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray',echo=FALSE}

# Estimate ARIMA models
model_1 <- arima(crude_oil_bx$diff_total_exports, order = c(1, 0, 0))
model_2 <- arima(crude_oil_bx$diff_total_exports, order = c(0, 0, 1))
model_3 <- arima(crude_oil_bx$diff_total_exports, order = c(1, 0, 1))

# Diagnostic testing
checkresiduals(model_1)
checkresiduals(model_2)
checkresiduals(model_3)
```

```{r, comparison of ARIMA,echo=FALSE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray',echo=FALSE}
# Compare models using information criteria
aic <- AIC(model_1, model_2, model_3)
bic <- BIC(model_1, model_2, model_3)

@@

The Prophet model has been selected with the following inputs:

    Period: 12 (indicating yearly seasonality)
    Order: 1 (considering the previous time point)
    Type: Additive (consistent magnitude of seasonality)

These inputs allow the model to capture the yearly seasonality in the data, incorporate the autoregressive nature of the series, and generate accurate forecasts by combining the trend and seasonal components.

```{r, prophet, echo=FALSE, results='markup',warning=FALSE,message=FALSE, fig.align='center', background= 'gray',echo=FALSE}
library(fable.prophet)

train_pr <- crude_oil_bx %>%
  filter(year(Date) <= 2021)

fit <- train_pr %>%
  model(
    arima = ARIMA(Total_Exports),
    ets = ETS(Total_Exports),
    additive_ETS=ETS(Total_Exports~error("A")+trend("A")+season("A")),
    prophet = prophet(Total_Exports~season(period=12,order=1,type="additive"))
  )

accuracy(fit)

fc <- fit %>%  forecast(h = 23)

fc %>% 
  autoplot(
    crude_oil_bx,
    level = NULL) +
  labs(
    y = "Amount of natural gas (BOE)",
    title = "Forecasts for U.S. Natural gas exports to OECD Europe") +
  guides(colour = guide_legend(title = "Forecast"))
```

After evaluating the accuracy of the models, it is evident that all models perform relatively similarly in terms of accuracy. However, there is a noticeable difference in the forecasted growth of crude oil exports. The Prophet Model, while providing a slightly more pessimistic forecast, predicts a lower growth rate compared to the ARIMA and ETS models. This suggests that the Prophet Model takes a more cautious approach in estimating the future growth of crude oil exports.

## Conclusion

In conclusion, this report compared the performance of ETS, ARIMA, and Prophet models with simple forecasting methods like Seasonal Naive and Drift. The analysis revealed that the advanced models, ETS, ARIMA, and Prophet, consistently outperformed the simple methods in terms of accuracy. These models captured the underlying patterns, seasonality, and trends in the crude oil export data more effectively. Additionally, the report highlighted the impact of geopolitical events, such as Russia's invasion of Ukraine, on the shifting export trends. Considering these findings, we believe that the observed trend in exports will likely continue in the future.

### 2023-05-22 report_Tomo-editinta.tex
updated

After conducting first order difference we see that data is now
stationary.

\includegraphics{report_Tomo-editinta_files/figure-latex/unnamed-chunk-1-1.pdf}
\includegraphics{report_Tomo-editinta_files/figure-latex/unnamed-chunk-1-2.pdf}

>>>>

\begin{verbatim}
## Warning: Removed 1 row containing missing values (`geom_line()`).
\end{verbatim}

\includegraphics{report_Tomo-editinta_files/figure-latex/first order difference-1.pdf}
After conducting first order difference we see that data is now
stationary.

\begin{verbatim}
## Warning: Removed 1 row containing missing values (`geom_line()`).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 1 rows containing missing values (`geom_point()`).
\end{verbatim}

\includegraphics{report_Tomo-editinta_files/figure-latex/unnamed-chunk-1-1.pdf}
